---
title: "Configuring a Crisis: Black Maternal Health and Bureaucratic Dispossession"
author: "Christian Hosam"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    number_sections: false

abstract: |
  To what extent does the unique structure of American healthcare perpetuate racial disparities in maternal mortality and morbidity rates? Maternal mortality in the United States is often configured as a crisis given the disproportionate number of maternal deaths in comparison to other advanced industrial countries. Much of the recent research and advocacy on combatting these disparities have focused on moving from individual frameworks of blame to centering stories of either provider-level implicit bias and/or racism as well as social determinants of health. One avenue that has been relatively unexplored is the structure of the American healthcare system, particularly the uniquely high proportion of healthcare functionality that is devolved to states and localities. Here, I look at the three counties within the state of Maryland with the highest concentration of Black residents to show how the types of health-care initiatives located at the county level focus on nudging birthing people into better outcomes in spite of medical and sociological literature showing that these interventions are misapplied.
  
bibliography: "https://api.citedrive.com/bib/02db2133-c341-42d0-8077-98b8e30b6c5d/references.bib?x=eyJpZCI6ICIwMmRiMjEzMy1jMzQxLTQyZDAtODA3Ny05OGI4ZTMwYjZjNWQiLCAidXNlciI6ICIzMTQ2IiwgInNpZ25hdHVyZSI6ICJjMjhkZmYzMjkzODM5YzNkYzJlZjc4YjE0ZGU4ZTMwYzYyNzEzNWY5NmQ4ZDUzNjNlMGZiZmIxZDQ0M2ZmMmFmIn0=/bibliography.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Package Import, include=FALSE}
library(devtools)
library(tidyverse)
library(usethis)
library(rticles)
library(bookdown)
library(kableExtra)
library(ggthemes)
```

# Introduction

That maternal mortality is at epidemic levels in the United States is not up for debate. The trends, both in terms of overall numbers (which spiked in the wake of the Covid-19 pandemic) and in the crisis's attendant racial disparities, have not budged in decades, even as there has been more coverage, more attention paid by federal policymakers, most notably with the formation of the Black Maternal Health Caucus in Congress (), and increased attention to structural factors by professional organizations such as the American Medical Association () and the Center for Disease Control.

The crisis thus becomes not just a matter of life-and-death for the birthing people caught in the crossfire, but also a matter of national embarrassment, given that the national average of 23.8 deaths per 100,000 live births dwarfs the country with the second highest rate, New Zealand, which has 13.6 deaths. The problem becomes especially appalling when added to the mix is the corollary statistic according to the CDC, over 84% of pregnancy-related deaths were preventable. But what about this crisis is so uniquely American? What features of American society and the American policy state imperils so many birthing people, and so many Black women in particular?

The extant literature, particularly within public health focuses on moving away from individualistic factors that serve to blame birthing people for their "risky" behaviors and instead move towards a framing that centers both social determinants of health () as well as in more recent years the racism that they face within the medical system as a result of provider-level racism, both in terms of mis- and under-diagnosis of pain and fear in addition to dismissal of their concerns in the immediate postpartum period. I affirm and agree with these fields. In this work, however, I intend to tackle the specificity of the American case by speaking more directly to a particular aspect of the American policy state that I argue also places Black birthing people in a unique form of risk: the federalist nature of American health care.

I will do this with three different approaches: the first is through the rise of community health programs within state and local governments. While the federal government often sets broad mandates for care within the United States, the authority for policies surrounding distribution of care and regulations about what care will look like in terms of what is necessary for birthing people to do under a program like Medicaid falls to the states and, increasingly, to local governments. This often means that to the extent that a birthing person is dealing with a government agent to assist with managing their pregnancy, this person is very rarely a federal official and is much more likely someone by employed by a county or city government. This might be benign on its face, but as I show, even as the country claims to be moving in a more progressive direction with respect to avoiding individualistic narratives, there is a disconnect between federal, state, and local frameworks for care, with the smaller governments often opting for more individualistic frameworks, serving to re-inscribe these cultural pathologies rather than undermining them.

The second is through the landscape of insurance coverage within the country. As Khiara Bridges vividly describes in her book *Reproducing Race,* being a recipient of Medicaid, which a disproportionate number of Black women are, places them within a particular context of surveillance by the state both in terms of the medical tests that they have to endure as well as the types of pathological narratives that they have to contend with being both poor and Black. By looking at the rates of change of those Black women who have private insurance (specifying between out-of-pocket and employer-covered insurance) as well as those who have public insurance (specifying between Medicaid and state insurance exchanges), I intend to show that while the type of surveillance that a birthing person will face might be subject to the type of health care they receive, the discrepancy is far lower and more complicated than we might intuitively imagine. I instead argue that even for those with private insurance, which is supposed to grant a level of autonomy and choice to the birthing person, the specter of race and anti-Blackness in particular can often mean that they are subject to many of the same types of pernicious pressures to subject themselves to increasingly invasive medical treatment that creates more and more opportunities for them to be judged as "unruly bodies" as well as become more likely to be seen as unfit and, as a result, be subject to state intervention of their bodies and of their children.

Finally, I look at OECD data to look at the ways in which the decomposition of healthcare expenditure in the United States is unique in many ways to its industrialized peers, shaping a healthcare system that allows ample opportunity for providers to play into their most base assumptions about all birthing people, but Black women in particular, because differences in funding between federal and state governments create an incentive-structure in which doctors have to make decisions that attend to both law as well as the interests of medical capital, putting them in the position of calculating between risk and care, and basing those decisions on their analysis of not just what is good for the birthing person but also what is most risk-neutral for their careers. This has the additional effect of structuring the two factors above. By devolving authority in complex, yet slipshod ways, the American healthcare system is something to be navigated for all people. For Black women in particular, they have to negotiate not only the mandates of whatever insurance coverage that they have in addition to whatever type of supplemental support that they might need from their local communities and governments but also be savvy enough to overcome the ever-present potential of racism and sexism affecting the type of medical treatment they receive. In my final section, I theorize on this phenomenon: calling it bureaucratic dispossession.

Taken together, the American maternal mortality crisis is one of racism and sexism that is faced most acutely by Black women. However, as I will show, these biases show up not just in the intimate spaces in which they are the most vulnerable but also in the very construction of how they receive care. In this way, Black women are yet again the canaries in the coal mine, highlighting the factors of the American political economy that lay right beneath the surface, imperiling us all but made clear through the specific and devastating impacts that they face.

# Entrenching the Technocratic Model of Pregnancy

# The Insurance Landscape for Black Birthing People in the United States

# American Exceptionalism and Healthcare Expenditure

# Theorizing Bureaucratic Dispossession
